Optimizing a PoS Tag Set for Dependency Parsing

### 2016-05-05 submission.tex
Minor changes, fixed references

\begin{table}
    \centering
    %\smaller[0.5]
    \begin{tabular}{@{}ll@{}}
        \toprule
        \textbf{Tag} & \textbf{Description} \\
        \midrule
        \texttt{adj} & Adjective \\
        \texttt{adv} & Adverb \\
        \texttt{det} & Determiner\\
        \texttt{inf-merke} & Infinitive marker \\
        \texttt{interj} & Interjection \\
        \texttt{konj} & Conjunction \\
        \texttt{prep} & Preposition \\
        \texttt{pron} & Pronoun \\
        \texttt{sbu} & Subordinate conjunction \\
        \texttt{subst} & Noun \\
        \texttt{ukjent} & Unknown (foreign word) \\
        \texttt{verb} & Verb \\
        \bottomrule
    \end{tabular}
    \caption{Overview of the PoS tag set of NDT.}
    \label{ndttagset}
\end{table}

@@

\begin{table}
    \centering
    %\smaller[0.5]
    \begin{tabular}{@{}ll@{}}
        \toprule
        \textbf{Head} & \textbf{Dependent} \\
        \midrule
        Preposition & Prepositional complement \\
        Finite verb & Complementizer \\
        First conjunct & Subsequent conjuncts \\
        Finite auxiliary & Lexical/main verb \\
        Noun & Determiner \\
        \bottomrule
    \end{tabular}
    \caption{Central head-dependent annotation choices in NDT.}
    \label{ndtannotation}
\end{table}

@@

\paragraph{Data Set Split}
As there was no standardized data set split of NDT due to its very recent
development, we needed to establish a data set split
(training/development/test). Our data set split of the treebank follows the
standard 80-10-10 (training/development/test) split and will be distributed
with the treebank and proposed as the new standard(?). In creating the data set
split, care has been taken to preserve contiguous texts in the various data
sets while keeping the split balanced in terms of genre. Our
proposed data set split was used in the Norwegian contribution to the Universal
Dependencies project \cite{Ovr:Hoh:16}.  The split will be made available at a
companion website.

@@

@@

\subsection{Baseline Experiments}
In an initial round of experiments, we concatenated the tag of each token with
its set of morphological features in order to map the original tag set to a
new, more fine-grained tag set (hereafter referred to as the \emph{full} tag
set). This resulted in a total of 368 tags, which is clearly very fine-grained.
The two initial tag sets, i.e., the original tag set comprising 19 tags and the
full tag set comprising 368 tags, thus represent two extremes in terms of
granularity. To evaluate the tag sets and investigate how the tag set
granularity would affect the performance of tagging and parsing, we trained and
evaluated TnT and Mate on the training and development data, respectively, on
the two tag sets. In Table \ref{inittagseteval}, we report the results of these
experiments. We see that the tagger accuracy drastically drops when going from
the original to the full tag set. TnT reports an accuracy of 97.47\% on the
original tag set, which is reduced to 93.48\% for the full tag set. These
results confirm our hypothesis that the very high linguistic quality in the
full, fine-grained tag set comes at the expense of drops in tagger performance.
However, the additional linguistic information provided by the full tag set
improves the parser performance. With the original tag set, Mate reports an LAS
of 87.01\% and a UAS of 90.19\%, which increases to 87.15\% and 90.39\%,
respectively, when using the full tag set. As we are looking for linguistically
informed distinctions that improve the syntactic parsing, these results are
promising and serve to indicate that additional morphological information
assists the syntactic parsers, motivating the optimization of the existing PoS
tag set.

@@

\subsection{Tag Set Experiments}
We modified the tags for nouns, verbs, adjectives, determiners and pronouns in
NDT by appending selected sets of morphological features to each tag in order
to increase the linguistic information expressed by the tags.  For each tag, we
first experimented with each of the features in isolation before employing
various combinations of them. We based our choices of combinations on how
promising the features are and what we deem worth investigating in terms of
linguistic utility, in order to see how the features might interact.

The morphological properties of the various parts-of-speech is reflected in the
morphological features associated with the respective PoS tags. For instance,
as nouns can take on gender, definiteness and number, additionally "being of"
genitive case, the treebank operates with features for gender, definiteness,
number and case with "accompanying" values. In addition to
morphological properties such as definiteness, tense and number, all classes
except for verbs can be distinguished "on" type, e.g., common nouns and proper
nouns, while pronouns can be reflexive, reciprocal, personal or interrogative.

\subsection{Experiments}
"Nearly" all tag set modifications leads to a drop in tagger accuracy, as
expected. Few exceptions... We observed increases in parser accuracy scores
with linguistically motivated tag set modifications for all PoS tags we
experimented with, the most improved being noun with an increase in LAS from
the original tag set by 1.80 percentage points.

\paragraph{Nouns}
In Norwegian, nouns are assigned gender (feminine, masculine, neuter),
definiteness (indefinite or definite) and number (singular or plural). There
is agreement in gender, definiteness and number between nouns and their
modifiers (adjectives and determiners). Additionally, NDT has a separate case
feature to distinguishing nouns in genitive case.

Experimenting with nouns, we found that all tag set modifications yielded large
increases in LAS and UAS. The most informative features are definiteness, which
leads to an increase in LAS by 1.26 percentage points, to 88.27\%, and type,
yielding an LAS of 88.07\%. Turning to combinations of features, we found that
the combination of type and case, as well as type and definiteness, were the
most promising, which led us to combine type, case and definiteness in a final
experiment, resulting in LAS of 88.81\% and UAS of 91.73\%, constituting large
increases from parsing with the original tag set, 1.80 percentage points
and 1.54 percentage points, respectively.

The results from tagging and parsing with modifications to nouns are reported
in Table \ref{substresults}. As nouns constitute the largest class in the
treebank by far (as shown in Table \ref{baselinetagerror}), the effects are
greatest for noun tokens in terms of overall change in performance. Apart from
case, none of the tag set modifications improves the tagging. However, they all
give rise to increases in parser accuracy scores. Genitive case marks
possession, hence nouns marked with genitive case are quite different from
other nouns, taking a noun phrase as complement. Distinguishing on type is
useful and informative, as evident by the presence of separate tags for proper
and common nouns in many tag sets, such as those of PTB, SUC and UD (described
in Section \ref{sec:extagsets}). When introducing the distinction of type, we
see a large increase in parser accuracy scores, with an LAS of 88.07\% and UAS
of 91.11\%, both exceeding the baseline by more than a percentage point.
Definiteness is the most informative feature for parsing, achieving LAS of
88.27\% and UAS of 91.42\%.

%\begin{table}
    %\centering
    %\smaller[0.5]
    %\begin{tabular}{@{}lllll@{}}
        %\toprule
        %\textbf{Feature(s)} & \textbf{MFT} & \textbf{Accuracy} &
        %\textbf{LAS} & \textbf{UAS} \\
        %\midrule
        %--- & \textbf{94.14\%} & 97.47\% & 87.01\% & 90.19\% \\
        %Case & 93.77\% & \textbf{97.48\%} & 87.63\% & 90.72\% \\
        %Definiteness & 89.67\% & 97.00\% & 88.27\% & 91.42\% \\
        %Gender & 89.54\% & 96.09\% & 87.21\% & 90.36\% \\
        %Number & 90.04\% & 96.37\% & 87.97\% & 91.00\% \\
        %Type & 91.90\% & 96.92\% & 88.07\% & 91.11\% \\
        %Case \& definiteness & 89.65\% & 97.03\% & 88.39\% & 91.44\% \\
        %Type \& case & 91.73\% & 96.92\% & 88.46\% & 91.51\% \\
        %Type \& definiteness & 89.65\% & 96.99\% & 88.44\% & 91.48\% \\
        %Type \& number & 90.02\% & 96.37\% & 87.95\% & 90.95\% \\
        %Type, case \& definiteness & 89.61\% & 97.05\% & \textbf{88.81\%} &
        %\textbf{91.73\%} \\
        %Type, definiteness \& number & 88.39\% & 96.46\% & 88.06\% & 91.12\%
        %\\
        %\bottomrule
    %\end{tabular}
    %\caption{Results of experiments with modified PoS tags for nouns.}
    %\label{substresults}
%\end{table}

\paragraph{Verbs}
Verbs are inflected for tense (infinitive, present, preterite, past perfect)
in Norwegian and can additionally take on mood (imperative, indicative) and
voice (active or passive). Note that voice and mood have only a single value,
\texttt{pass} (passive) and \texttt{imp} (imperative), respectively. Verbs
which are not passive are implicitly active, and verbs which are not imperative
are indicative.

Introducing features in isolation, mood is the only feature leading to
increases in LAS, with a reported LAS of 87.04\%. Imperative clauses are
fundamentally different from indicative clauses, as they lack an overt subject.
Tense yields an LAS of 86.97\% while voice yields an LAS of 86.96\%. Combining
mood and tense resulted in an LAS of 87.12\% and UAS of 90.31\%.

In an additional experiment, we mapped the verb tenses (mood, in the case of
imperative) to finiteness.  All verbs have finiteness, hence this distinction
has broad coverage.  This mapping is syntactically grounded as finite verbs and
nonfinite verbs appear in completely different syntactic construction, and
proved to greatly improve the parsing, as we saw the overall largest parser
accuracy scores, with 87.30\% for LAS and 90.43\% for UAS, 0.29 and 0.24
percentage points higher than the baseline, respectively. This coincides with
the observations seen for Swedish in \newcite{Ovr:08}, where finiteness was found
to be a very beneficial linguistic feature for parsing.

%\begin{table}
    %\centering
    %\smaller[0.5]
    %\begin{tabular}{@{}lllll@{}}
        %\toprule
        %\textbf{Feature(s)} & \textbf{MFT} & \textbf{Accuracy} &
        %\textbf{LAS} & \textbf{UAS} \\
        %\midrule
        %--- & \textbf{94.14\%} & \textbf{97.47\%} & 87.01\% & 90.19\% \\
        %Mood & 94.12\% & 97.43\% & 87.04\% & 90.19\% \\
        %Tense & 93.74\% & 97.30\% & 86.97\% & 90.18\% \\
        %Voice & 94.13\% & 97.45\% & 86.96\% & 90.09\% \\
        %Mood \& tense & 93.74\% & 97.31\% & 87.12\% & 90.31\% \\
        %Voice \& tense & 93.74\% & 97.28\% & 86.99\% & 90.15\% \\
        %Mood, tense \& voice & 93.74\% & 97.27\% & 86.83\% & 90.05\% \\
        %Finiteness & 93.72\% & 97.35\% & \textbf{87.30\%} &
        %\textbf{90.43\%} \\
        %\bottomrule
    %\end{tabular}
    %\caption{Results of experiments with modified PoS tags for verbs.}
    %\label{verbresults}
%\end{table}

\paragraph{Adjectives}
Adjectives agree with the noun they modify in terms of gender, number and
definiteness in Norwegian. Adjectives are also inflected for degree, either
positive, comparative or superlative.

All features except for number (LAS of 86.99\%) led to increases in parser
accuracy scores. Degree being the most promising with a reported LAS of
87.29\%, while distinguishing adjectives on definiteness yield an LAS of
87.14\% and gender leads to LAS of 87.10\%.

Combining the most promising features, we found that none of the combinations
surpass the LAS with the distinction of degree alone. Definiteness and degree
with an LAS of 87.23\% and definiteness and number with an LAS of 87.27\% and
UAS identical to that of degree alone. The most fine-grained distinction,
definiteness, degree and number, yields an LAS of 87.14\%.

%\begin{table}
    %\centering
    %\smaller[0.5]
    %\begin{tabular}{@{}lllll@{}}
        %\toprule
        %\textbf{Feature(s)} & \textbf{MFT} & \textbf{Accuracy} &
        %\textbf{LAS} & \textbf{UAS} \\
        %\midrule
        %--- & \textbf{94.14\%} & \textbf{97.47\%} & 87.01\% & 90.19\% \\
        %Definiteness & 93.45\% & 96.84\% & 87.14\% & 90.29\% \\
        %Degree & 94.13\% & 97.41\% & \textbf{87.29\%} & \textbf{90.44\%} \\
        %Gender & 93.56\% & 96.89\% & 87.10\% & 90.25\% \\
        %Number & 93.51\% & 96.71\% & 86.99\% & 90.10\% \\
        %Type & 94.12\% & 97.40\% & 87.11\% & 90.25\% \\
        %Definiteness \& degree & 93.45\% & 96.81\% & 87.23\% & 90.39\% \\
        %Definiteness \& gender & 92.94\% & 96.31\% & 87.18\% & 90.39\% \\
        %Definiteness \& number & 93.48\% & 96.78\% & 87.27\% &
        %\textbf{90.44\%} \\
        %Degree \& gender & 93.56\% & 96.87\% & 87.00\% & 90.16\% \\
        %Degree \& number & 93.49\% & 96.76\% & 87.13\% & 90.26\% \\
        %Definiteness, degree \& number & 93.47\% & 96.81\% & 87.14\% & 90.30\% \\
        %\bottomrule
    %\end{tabular}
    %\caption{Results of experiments with modified PoS tags for adjectives.}
    %\label{adjresults}
%\end{table}

@@

The results from the experiments with determiners are shown in Table
\ref{detresults}.  Introducing the type led to an increase in tagger accuracy
by 0.14 percentage points to 97.61\%, while marginally impacting the parsing,
with LAS of 87.00\%, 0.01 percentage points below the baseline, and UAS of
90.11\%, 0.08 percentage points below the baseline.
%The fact that the baseline
%remains the same is especially worth noting, as increased tag set granularity
%almost inevitably leads to a drop in the MFT baseline.
The increase in tagger
accuracy when introducing the distinction of type is noteworthy, as we expected
the finer granularity to lead to a decrease in accuracy. This serves to
indicate that more fine-grained distinctions for determiners, which is a quite
disparate category in the treebank, may be quite useful for tagging. However,
as it has negative impact on the syntactic parsing, we can conclude that the
type of a determiner does not assist in generalizing syntactic patterns, as
most determiners, regardless of type, appear in the same syntactic
constructions (i.e., before a noun or noun phrase).

Gender, on the other hand, improved the parsing, but complicated the tagging,
as the various genders are often difficult to differentiate, especially so in
the case of masculine and feminine, which share many of the same determiners.
The number of a determiner, i.e., singular or plural, led to a small increase
in tagger accuracy and LAS, while marginally lower UAS, 90.18\%, 0.01
percentage points lower than that of the original tag set. Almost all
determiners have number, and the introduction of this distinction led to small
increases in tagger accuracy and LAS, but marginally lower UAS. The
introduction of definiteness to the determiners led to the best parsing
results, LAS of 87.30\% and UAS of 90.42\%, while also increasing the tagger
accuracy slightly. The increase in LAS and UAS is rather interesting, as there
are only 121 "determiner tokens" with marked definiteness in the development
data. As this accounts for a very small number of tokens, coupled with the
previously noted considerations, we did not consider further fine-grained
modifications with definiteness. This goes to show that tokens with overt
definiteness have noticeable impact on the syntactic parsing, and that
distinguishing on definiteness is very beneficial.

%\begin{table}
    %\centering
    %\smaller[0.5]
    %\begin{tabular}{@{}lllll@{}}
        %\toprule
        %\textbf{Feature} & \textbf{MFT} & \textbf{Accuracy} &
        %\textbf{LAS} & \textbf{UAS} \\
        %\midrule
        %--- & \textbf{94.14\%} & 97.47\% & 87.01\% & 90.19\% \\
        %Definiteness & 94.13\% & 97.49\% & \textbf{87.30\%} &
        %\textbf{90.42\%} \\
        %Gender & 93.86\% & 97.28\% & 87.09\% & 90.31\% \\
        %Number & 94.06\% & 97.49\% & 87.04\% & 90.18\% \\
        %Type & 94.14\% & \textbf{97.61\%} & 87.00\% & 90.11\% \\
        %\bottomrule
    %\end{tabular}
    %\caption{Results of experiments with modified PoS tags for determiners.}
    %\label{detresults}
%\end{table}

@@

%\begin{table}
    %\centering
    %\smaller[0.5]
    %\begin{tabular}{@{}lllll@{}}
        %\toprule
        %\textbf{Feature(s)} & \textbf{MFT} & \textbf{Accuracy} &
        %\textbf{LAS} & \textbf{UAS} \\
        %\midrule
        %--- & \textbf{94.14\%} & 97.47\% & 87.01\% & 90.19\% \\
        %Case & 94.12\% & 97.50\% & 87.08\% & 90.21\% \\
        %Gender & 94.13\% & 97.48\% & 87.06\% & 90.23\% \\
        %Number & 94.13\% & 97.49\% & 87.21\% & 90.33\% \\
        %Person & 94.14\% & 97.49\% & 87.22\% & 90.32\% \\
        %Type & 94.14\% & 97.48\% & 87.19\% & 90.40\% \\
        %Number \& person & 94.13\% & 97.49\% & 96.98\% & 90.16\% \\
        %Type \& case & 94.12\% & 97.51\% & \textbf{87.30\%} &
        %\textbf{90.41\%} \\
        %Type \& number & 94.13\% & 97.49\% & 87.27\% & \textbf{90.41\%} \\
        %Type \& person & 94.14\% & 97.49\% & 87.00\% & 90.14\% \\
        %Type, case \& number & 94.12\% & \textbf{97.52\%} & 87.11\% & 90.36\% \\
        %\bottomrule
    %\end{tabular}
    %\caption{Results of experiments with modified PoS tags for pronouns.}
    %\label{pronresults}
%\end{table}

@@

\begin{table}
    \centering
    \smaller[0.5]
    \begin{tabular}{@{}llllll@{}}
        \toprule
        \textbf{Data Set} & \textbf{Tag Set} & \textbf{MFT} & \textbf{Accuracy}
        & \textbf{LAS} & \textbf{UAS} \\
        \midrule
        \multirow{2}{*}{Dev}
        & Original & \textbf{94.14\%} & \textbf{97.47\%} & 87.01\% & 90.19\% \\
        %Full & 85.15\% & 93.48\% & 87.15\% & 90.39\% \\
        & Optimized & 89.20\% & 96.85\% & \textbf{88.87\%} & \textbf{91.78\%} \\
        \midrule
        \multirow{2}{*}{Test}
        & Original & \textbf{94.22\%} & \textbf{97.38\%} & 86.64\% & 90.07\% \\
        %& Full & 84.27\% & 92.79\% & 87.03\% & 90.16\% \\
        & Optimized & 88.08\% & 96.35\% & \textbf{88.55\%} & \textbf{91.41\%} \\
        \bottomrule
    \end{tabular}
    \caption{Results of tagging and parsing with the optimized tag set,
        compared to the initial tag sets. Trained and tested using
        automatically assigned tags from TnT.}
    \label{finalresults}
\end{table}

>>>>

\begin{table}
    \centering
    \smaller[0.5]
    \begin{tabular}{@{}ll@{}}
        \toprule
        \textbf{Tag} & \textbf{Description} \\
        \midrule
        \texttt{adj} & Adjective \\
        \texttt{adv} & Adverb \\
        \texttt{det} & Determiner\\
        \texttt{inf-merke} & Infinitive marker \\
        \texttt{interj} & Interjection \\
        \texttt{konj} & Conjunction \\
        \texttt{prep} & Preposition \\
        \texttt{pron} & Pronoun \\
        \texttt{sbu} & Subordinate conjunction \\
        \texttt{subst} & Noun \\
        \texttt{ukjent} & Unknown (foreign word) \\
        \texttt{verb} & Verb \\
        \bottomrule
    \end{tabular}
    \caption{Overview of the PoS tag set of NDT.}
    \label{ndttagset}
\end{table}

@@

\begin{table}
    \centering
    \smaller[0.5]
    \begin{tabular}{@{}ll@{}}
        \toprule
        \textbf{Head} & \textbf{Dependent} \\
        \midrule
        Preposition & Prepositional complement \\
        Finite verb & Complementizer \\
        First conjunct & Subsequent conjuncts \\
        Finite auxiliary & Lexical/main verb \\
        Noun & Determiner \\
        \bottomrule
    \end{tabular}
    \caption{Central head-dependent annotation choices in NDT.}
    \label{ndtannotation}
\end{table}

@@

\paragraph{Data Set Split}
As there was no standardized data set split of NDT due to its very recent
development, we needed to establish a data set split
(training/development/test). Our data set split of the treebank follows the
standard 80-10-10 (training/development/test) split and will be distributed
with the treebank and proposed as the new standard(?). In creating the data set
split, care has been taken to preserve contiguous texts in the various data
sets while keeping the split balanced in terms of genre. Our
proposed data set split was used in the Norwegian contribution to the Universal
Dependencies project \cite{Ovr:Hoh:16}. The split will be made available at a
companion website. In our experiments, we train on the training data and
evaluate on the development data, before finally evaluating our optimized tag
set on the held-out test data set.

@@

\paragraph{Evaluation}
To evaluate tagging and parsing with the various tag set modifications in our
experiments, we employed commonly used evaluation metrics. Tagging is evaluated
in terms of accuracy (reported by the TnT-included \texttt{tnt-diff} script),
while labeled attachment score (LAS) and unlabeled attachment score (UAS) are
computed by the
\texttt{eval.pl}\footnote{\url{http://ilk.uvt.nl/conll/software.html}} script
(used in the CoNLL shared tasks) to evaluate the parsing.

@@

\subsection{Baseline Experiments}
In an initial round of experiments, we concatenated the tag of each token with
its set of morphological features in order to map the original tag set to a
new, more fine-grained tag set (hereafter referred to as the \emph{full} tag
set). This resulted in a total of 368 tags, which is clearly very fine-grained.
The two initial tag sets, i.e., the original tag set comprising 19 tags and the
full tag set comprising 368 tags, thus represent two extremes in terms of
granularity. To evaluate the initial tag sets and investigate how the tag set
granularity would affect the performance of tagging and parsing, we ran
baseline experiments with the two tag sets. In Table \ref{inittagseteval}, we
report the results of these experiments. We see that the tagger accuracy
drastically drops when going from the original to the full tag set. TnT reports
an accuracy of 97.47\% on the original tag set, which is reduced to 93.48\% for
the full tag set. These results confirm our hypothesis that the very high
linguistic quality in the full, fine-grained tag set comes at the expense of
drops in tagger performance.  However, the additional linguistic information
provided by the full tag set improves the parser performance. With the original
tag set, Mate reports an LAS of 87.01\% and a UAS of 90.19\%, which increases
to 87.15\% and 90.39\%, respectively, using the full tag set. As we are looking
for linguistically informed distinctions that improve the syntactic parsing,
these results are promising and serve to indicate that additional linguistic
information assists the syntactic parsers, motivating the optimization of the
existing PoS tag set.

@@

\subsection{Tag Set Experiments}
We modified the tags for nouns (\texttt{subst}), verbs (\texttt{verb}),
adjectives (\texttt{adj}), determiners (\texttt{det}) and pronouns
(\texttt{pron}) in NDT by appending selected sets of morphological features to
each tag in order to increase the linguistic information expressed by the tags.
For each tag, we first experimented with each of the available features in
isolation before employing various combinations of them. We based our choices
of combinations on how promising the features are and what we deem worth
investigating in terms of linguistic utility, in order to see how the features
might interact.

The morphological properties of the various parts-of-speech are reflected in
the morphological features associated with the respective PoS tags. For
instance, as nouns in Norwegian can take on gender, definiteness and number,
additionally marked with genitive case, the treebank operates with features for
gender, definiteness, number and case.
%with "accompanying" values
In addition to morphological properties such as definiteness, tense and number,
all classes except for verbs can be distinguished "on" type, e.g., common nouns
and proper nouns, while pronouns can be reflexive, reciprocal, personal or
interrogative.

%"Nearly" all tag set modifications leads to a drop in tagger accuracy, as
%expected. Few exceptions... We observed increases in parser accuracy scores
%with linguistically motivated tag set modifications for all PoS tags we
%experimented with, the most improved being noun with an increase in LAS from
%the original tag set by 1.80 percentage points.

\paragraph{Nouns}
In Norwegian, nouns are assigned gender (feminine, masculine or neuter),
definiteness (definite or indefinite) and number (singular or plural). There
is agreement in gender, definiteness and number between nouns and their
modifiers (adjectives and determiners). Additionally, NDT has a separate case
feature for distinguishing nouns in genitive case.

As nouns constitute the largest class in the treebank by far, the effects are
greatest for noun tokens in terms of overall change in performance. The results
of experimenting with tag set modifications for nouns are shown in Table
\ref{substresults}, and reveal that, apart from case, none of the tag set
modifications improves the tagging. However, they all give rise to increases in
parser accuracy scores. Genitive case marks possession, hence nouns marked with
genitive case are quite different from other nouns, taking a noun phrase as
complement. Distinguishing on type is useful and informative, as evident by the
presence of separate tags for proper and common nouns in many tag sets, such as
those of Penn Treebank and Stockholm-Umeå corpus.

The most informative features are definiteness, which
leads to an increase in LAS by 1.26 percentage points to 88.27\%, and type,
yielding an LAS of 88.07\%. Turning to combinations of features, we found that
the combination of type and case, as well as type and definiteness, were the
most promising, which led us to combine type, case and definiteness in a final
experiment, resulting in LAS of 88.81\% and UAS of 91.73\%, constituting large
increases from parsing with the original tag set, by 1.80 percentage points
and 1.54 percentage points, respectively.

\begin{table}
    \centering
    \smaller[1]
    \begin{tabular}{@{}llll@{}}
        \toprule
        \textbf{Feature(s)} & \textbf{Acc} & \textbf{LAS} & \textbf{UAS}
        \\
        \midrule
        --- & 97.47\% & 87.01\% & 90.19\% \\
        Case & \textbf{97.48\%} & 87.63\% & 90.72\% \\
        Definiteness & 97.00\% & 88.27\% & 91.42\% \\
        Gender & 96.09\% & 87.21\% & 90.36\% \\
        Number & 96.37\% & 87.97\% & 91.00\% \\
        Type & 96.92\% & 88.07\% & 91.11\% \\
        Case \& definiteness & 97.03\% & 88.39\% & 91.44\% \\
        Type \& case & 96.92\% & 88.46\% & 91.51\% \\
        Type \& definiteness & 96.99\% & 88.44\% & 91.48\% \\
        Type \& number & 96.37\% & 87.95\% & 90.95\% \\
        Type, case \& definiteness & 97.05\% & \textbf{88.81\%} &
        \textbf{91.73\%} \\
        %Type, definiteness \& number & 96.46\% & 88.06\% & 91.12\% \\
        \bottomrule
    \end{tabular}
    \caption{Results of experiments with modified PoS tags for nouns.}
    \label{substresults}
\end{table}

\paragraph{Verbs}
Verbs are inflected for tense (infinitive, present, preterite, past perfect) in
Norwegian and additionally exhibit mood (imperative, indicative) and voice
(active or passive). Note that both voice and mood have only a single value in
the treebank, \texttt{pass} (passive) and \texttt{imp} (imperative),
respectively.  Verbs which are not passive are implicitly active, and verbs
which are not imperative are indicative.

Table \ref{verbresults} presents the results from tagging and parsing with
modified verb tags. Introducing features in isolation, mood is the only feature
leading to an increase in LAS, with a reported LAS of 87.04\%. Imperative
clauses are fundamentally different from indicative clauses, as they lack an
overt subject.  Although voice is a very distinguishing property for verbs, and
passive verbs appear in clauses very different from those with active verbs,
introducing this distinction in the tag set leads to a drop in LAS of 0.05
percentage points.  Distinguishing between the various tenses yields an LAS of
86.97\%. Combining the two most promising features, mood and tense, resulted in
an LAS of 87.12\% and UAS of 90.31\%.

In an additional experiment, we mapped the verb tenses (and mood, in the case
of imperative) to finiteness. All verbs have finiteness, hence this distinction
has broad coverage. This mapping is syntactically grounded as finite verbs and
nonfinite verbs appear in completely different syntactic constructions, and
proved to greatly improve the parsing, as we saw the overall largest parser
accuracy scores, with 87.30\% for LAS and 90.43\% for UAS, 0.29 and 0.24
percentage points higher than the baseline, respectively. This coincides with
the observations seen for Swedish in \newcite{Ovr:08}, where finiteness was
found to be a very beneficial linguistic feature for parsing.

\begin{table}
    \centering
    \smaller[0.5]
    \begin{tabular}{@{}llll@{}}
        \toprule
        \textbf{Feature(s)} & \textbf{Acc} & \textbf{LAS} & \textbf{UAS} \\
        \midrule
        --- & \textbf{97.47\%} & 87.01\% & 90.19\% \\
        Mood & 97.43\% & 87.04\% & 90.19\% \\
        Tense & 97.30\% & 86.97\% & 90.18\% \\
        Voice & 97.45\% & 86.96\% & 90.09\% \\
        Mood \& tense & 97.31\% & 87.12\% & 90.31\% \\
        Voice \& tense & 97.28\% & 86.99\% & 90.15\% \\
        Mood, tense \& voice & 97.27\% & 86.83\% & 90.05\% \\
        Finiteness & 97.35\% & \textbf{87.30\%} & \textbf{90.43\%} \\
        \bottomrule
    \end{tabular}
    \caption{Results of experiments with modified PoS tags for verbs.}
    \label{verbresults}
\end{table}

\paragraph{Adjectives}
Adjectives agree with the noun they modify in terms of gender, number and
definiteness in Norwegian. Adjectives are also inflected for degree (positive,
comparative or superlative).

Table \ref{adjresults}

All features except for number led to increases in parser accuracy scores.
Degree being the most promising with a reported LAS of 87.29\%, while
distinguishing adjectives on definiteness yield an LAS of 87.14\% and gender
leads to LAS of 87.10\%.

Combining the most promising features, we found that none of the combinations
surpass the LAS with the distinction of degree alone. Definiteness and degree
with an LAS of 87.23\% and definiteness and number with an LAS of 87.27\% and
UAS identical to that of degree alone.

Turning to combinations of features, definiteness and number achieve the best
results, very close to that of degree, with 0.02 percentage points lower LAS
and identical UAS. Adjectives agree with their head noun and determiner in
definiteness and number, making this an expected improvement. The combination
of definiteness and degree is also quite promising, obtaining LAS of 87.23\%
and UAS of 90.39\%. It is interesting that none of the combinations surpass the
experiment with degree alone, which indicates that degree does not interact
with the other features in any syntactically significant way.

\begin{table}
    \centering
    \smaller[0.5]
    \begin{tabular}{@{}llll@{}}
        \toprule
        \textbf{Feature(s)} & \textbf{Acc} & \textbf{LAS} & \textbf{UAS}
        \\
        \midrule
        --- & \textbf{97.47\%} & 87.01\% & 90.19\% \\
        Definiteness & 96.84\% & 87.14\% & 90.29\% \\
        Degree & 97.41\% & \textbf{87.29\%} & \textbf{90.44\%} \\
        Gender & 96.89\% & 87.10\% & 90.25\% \\
        Number & 96.71\% & 86.99\% & 90.10\% \\
        Type & 97.40\% & 87.11\% & 90.25\% \\
        Definiteness \& degree & 96.81\% & 87.23\% & 90.39\% \\
        Definiteness \& gender & 96.31\% & 87.18\% & 90.39\% \\
        Definiteness \& number & 96.78\% & 87.27\% & \textbf{90.44\%} \\
        Degree \& gender & 96.87\% & 87.00\% & 90.16\% \\
        Degree \& number & 96.76\% & 87.13\% & 90.26\% \\
        %Definiteness, degree \& number & 96.81\% & 87.14\% & 90.30\% \\
        \bottomrule
    \end{tabular}
    \caption{Results of experiments with modified PoS tags for adjectives.}
    \label{adjresults}
\end{table}

@@

The results from the experiments with determiners are shown in Table
\ref{detresults}.  Introducing the distinction of type led to an increase in
tagger accuracy by 0.14 percentage points to 97.61\%, while marginally
impacting the parsing, with LAS of 87.00\%, 0.01 percentage points below that
of the original tag set.  The increase in tagger accuracy when introducing the
distinction of type is noteworthy, as we expected the finer granularity to lead
to a decrease in accuracy. This serves to indicate that more fine-grained
distinctions for determiners, which is a quite disparate category in the
treebank, may be quite useful for tagging. However, as it has negative impact
on the syntactic parsing, we can conclude that the type of a determiner does
not assist in generalizing syntactic patterns, as most determiners, regardless
of type, appear in the same syntactic constructions (i.e., before a noun or
noun phrase).

Gender, on the other hand, improved the parsing, but complicated the tagging,
as the various genders are often difficult to differentiate, especially so in
the case of masculine and feminine, which share many of the same determiners.
The number of a determiner, i.e., singular or plural, led to a small increase
in tagger accuracy and LAS, while marginally lower UAS, 90.18\%, 0.01
percentage points lower than that of the original tag set. The
introduction of definiteness to the determiners led to the best parsing
results, LAS of 87.30\% and UAS of 90.42\%, while also increasing the tagger
accuracy slightly. The increase in LAS and UAS is rather interesting, as there
are only 121 "determiner tokens" with marked definiteness in the development
data. As this accounts for a very small number of tokens, coupled with the
previously noted considerations, we did not consider further fine-grained
modifications with definiteness. This goes to show that tokens with overt
definiteness have noticeable impact on the syntactic parsing, and that
distinguishing on definiteness is very beneficial.

\begin{table}
    \centering
    \smaller[0.5]
    \begin{tabular}{@{}llll@{}}
        \toprule
        \textbf{Feature} & \textbf{Acc} & \textbf{LAS} & \textbf{UAS} \\
        \midrule
        --- & 97.47\% & 87.01\% & 90.19\% \\
        Definiteness & 97.49\% & \textbf{87.30\%} & \textbf{90.42\%} \\
        Gender & 97.28\% & 87.09\% & 90.31\% \\
        Number & 97.49\% & 87.04\% & 90.18\% \\
        Type & \textbf{97.61\%} & 87.00\% & 90.11\% \\
        \bottomrule
    \end{tabular}
    \caption{Results of experiments with modified PoS tags for determiners.}
    \label{detresults}
\end{table}

@@

\begin{table}
    \centering
    \smaller[0.5]
    \begin{tabular}{@{}llll@{}}
        \toprule
        \textbf{Feature(s)} & \textbf{Acc} & \textbf{LAS} & \textbf{UAS} \\
        \midrule
        --- & 97.47\% & 87.01\% & 90.19\% \\
        Case & 97.50\% & 87.08\% & 90.21\% \\
        Gender & 97.48\% & 87.06\% & 90.23\% \\
        Number & 97.49\% & 87.21\% & 90.33\% \\
        Person & 97.49\% & 87.22\% & 90.32\% \\
        Type & 97.48\% & 87.19\% & 90.40\% \\
        Number \& person & 97.49\% & 96.98\% & 90.16\% \\
        Type \& case & 97.51\% & \textbf{87.30\%} & \textbf{90.41\%} \\
        Type \& number & 97.49\% & 87.27\% & \textbf{90.41\%} \\
        Type \& person & 97.49\% & 87.00\% & 90.14\% \\
        Type, case \& number & \textbf{97.52\%} & 87.11\% & 90.36\% \\
        \bottomrule
    \end{tabular}
    \caption{Results of experiments with modified PoS tags for pronouns.}
    \label{pronresults}
\end{table}

@@

\begin{table*}
    \centering
    \smaller[0.5]
    \begin{tabular}{@{}llllll@{}}
        \toprule
        \textbf{Data Set} & \textbf{Tag Set} & \textbf{MFT} & \textbf{Accuracy}
        & \textbf{LAS} & \textbf{UAS} \\
        \midrule
        \multirow{2}{*}{Dev}
        & Original & \textbf{94.14\%} & \textbf{97.47\%} & 87.01\% & 90.19\% \\
        %Full & 85.15\% & 93.48\% & 87.15\% & 90.39\% \\
        & Optimized & 89.20\% & 96.85\% & \textbf{88.87\%} & \textbf{91.78\%} \\
        \midrule
        \multirow{2}{*}{Test}
        & Original & \textbf{94.22\%} & \textbf{97.38\%} & 86.64\% & 90.07\% \\
        %& Full & 84.27\% & 92.79\% & 87.03\% & 90.16\% \\
        & Optimized & 88.08\% & 96.35\% & \textbf{88.55\%} & \textbf{91.41\%} \\
        \bottomrule
    \end{tabular}
    \caption{Results of tagging and parsing with the optimized tag set,
        compared to the initial tag sets. Trained and tested using
        automatically assigned tags from TnT.}
    \label{finalresults}
\end{table*}
